---
title: "02_differential accessibility"
output: html_document
date: "2026-06-23"
---



In the first part of the analysis, we identify genomic regions whose chromatin accessibility differs between BV2 microglial cells co-cultured with glioma cells under hypoxia versus normoxia. Differentially accessible peaks often mark regulatory regions (promoters, enhancers) that are dynamically remodeled in response to hypoxic stress, although accessibility changes alone do not prove that a region is actively driving transcription. They can therefore be used as a starting point for downstream functional and motif enrichment analyses (Parts 3–4), and can be compared with histone modification data to support or refine interpretation of regulatory activity (Part 5).


1. Define the sample sheet
2. Create consensus peak set
3. Count ATAC-seq reads from each BAM file over the consensus peaks
4. Normalize read counts across samples (TMM) and filter low-count peaks
5. Perform differential accessibility testing 

## Setup
```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(SummarizedExperiment)
  library(edgeR) 
  library(epiwraps)
  library(ggplot2)
  library(sechm) 
  library(GenomicAlignments)
  library(Rsamtools)
  library(SummarizedExperiment)
})
```




## Define the sample sheet

Sample metadata were first inspected by loading the SraRunTable_hypoxia_ATAC.csv file, which contains the original SRA run information (accessions, experimental conditions, etc.) downloaded for this dataset. This table was used to verify the correct sample identifiers and conditions before manually constructing the samples data frame used in the downstream analysis.


```{r}

ATAC_CSV <- read.csv("data/ATAC/SraRunTable_hypoxia_ATAC.csv")
ATAC_CSV

```

```{r}
# Manually define the sample sheet: GEO/SRA accessions, matched to experimental condition (hypoxia vs normoxia) and biological replicate

samples <- data.frame(
  gsm       = c("GSM8574835","GSM8574836","GSM8574833","GSM8574834"),
  srr       = c("SRR31003493","SRR31003492","SRR31003495","SRR31003494"),
  condition = c("hypoxia","hypoxia","normoxia","normoxia"),
  rep       = c(1, 2, 1, 2),
  stringsAsFactors = FALSE
)
```


## Create consensus peak set

To compare accessibility across samples at the same genomic regions, we first build a single consensus peak set by merging all per-sample peaks into one common set of intervals.

```{r}
# Build paths to the per-sample narrowPeak files and import them as GRanges

peaks <- file.path("./data/ATAC_narrowPeak",
  paste0(samples$gsm, "_BV2_cc_",
         ifelse(samples$condition == "hypoxia", "H", "N"),
         "_n", samples$rep, ".narrowPeak.gz"))
names(peaks) <- samples$srr

peaks_list <- lapply(peaks, rtracklayer::import)



# Merge (union) all per-sample peaks into a single consensus peak set, collapsing overlapping intervals into one region

merged_peaks <- reduce(unlist(GRangesList(peaks_list)))
cat("Total consensus peaks:", length(merged_peaks), "\n")

```

## Count ATAC-seq reads from each BAM file over the consensus peaks


We take the raw aligned reads (BAM files) and count, for each sample, how many reads fall into each consensus peak. This produces a peaks-by-samples count matrix, which is the standard input format required for differential analysis with edgeR.

```{r}
bam_files <- file.path("./data/ATAC", paste0(samples$srr, ".bam"))
names(bam_files) <- samples$srr
```



```{r}

# Count paired-end ATAC-seq reads from each BAM file overlapping the consensus peak set, producing a peaks-by-samples count matrix stored as a SummarizedExperiment object (se)

bam_list <- BamFileList(bam_files, yieldSize = 2000000)

se <- summarizeOverlaps(
  features = merged_peaks,
  reads = bam_list,
  mode = "Union",
  singleEnd = FALSE,
  ignore.strand = TRUE
)

colnames(se) <- samples$srr

colData(se)$condition <- samples$condition
colData(se)$rep <- samples$rep
```

```{r}
saveRDS(se, file = "data/peakCounts.rds")
```



# Standard differential analysis with edgeR and TMM normalization

The raw read counts cannot be compared directly between samples, since they reflect both true biological signal and technical factors such as sequencing depth. Here we use edgeR's standard workflow to normalize the counts, remove uninformative peaks, and statistically test each peak for a significant accessibility difference between hypoxia and normoxia.


```{r}
# Re-attach metadata explicitly in case colData was lost on reload
se <- readRDS("data/peakCounts.rds")

colData(se)$condition <- c("hypoxia", "hypoxia", "normoxia", "normoxia")
colData(se)$rep       <- c(1, 2, 1, 2)

dds <- edgeR::DGEList(
  counts = assay(se),
  group  = factor(colData(se)$condition, levels = c("normoxia", "hypoxia"))
)
# Build an edgeR DGEList from the raw peak counts, grouped by condition

dds <- edgeR::DGEList(
  counts = assay(se),
  group  = factor(se$condition, levels = c("normoxia", "hypoxia"))
)

# TMM normalization accounts for differences in sequencing depth and library composition between samples

dds <- edgeR::calcNormFactors(dds)


# Remove low-count peaks that do not carry enough statistical power for reliable differential testing

keep <- edgeR::filterByExpr(dds)
cat("\nPeaks kept after low-count filter:", sum(keep), "/", nrow(dds), "\n")
dds <- dds[keep, ]


# Estimate dispersion and perform an exact test for the normoxia vs. hypoxia comparison

dds <- edgeR::estimateDisp(dds)
et  <- exactTest(dds, pair = c("normoxia", "hypoxia"))
res <- as.data.frame(topTags(et, n = Inf))
```

We use TMM because, compared to other normalization methods like simple library-size scaling, it automatically identifies and normalizes against the stable, non-differential majority of peaks rather than relying on total read counts alone — avoiding bias from the strongly hypoxia-responsive peaks without requiring us to manually define a reference set.

#results

```{r}
dir.create("results_da", showWarnings = FALSE, recursive = TRUE)
saveRDS(list(
  res = res,   
  se  = se     
), file = file.path("results_da", "part1_results.rds"))
```

```{r, fig.cap="Volcano plot of differential chromatin accessibility between hypoxic and normoxic BV2-glioma co-cultures."}
head(res)

ggplot(res, aes(logFC, -log10(FDR))) +
  geom_point() +
  ggtitle("Differential chromatin accessibility: hypoxia vs. normoxia") +
  theme(plot.title = element_text(hjust = 0.5))
```
The logFC shows the direction and magnitude of the accessibility change — points to the right are more accessible in hypoxia, points to the left are more accessible in normoxia. The -log10 FDR shows statistical significance, so the higher a point sits, the more confident we are that the change is real rather than noise.

 
`
```{r}
cat("More open in hypoxia: ",  sum(res$FDR < 0.05 & res$logFC >  1, na.rm=TRUE), "\n")
cat("More open in normoxia:", sum(res$FDR < 0.05 & res$logFC < -1, na.rm=TRUE), "\n")
```


## Interpretation and Discussion

The plot shows widespread, fairly symmetric changes in both directions. With FDR < 0.05 and |logFC| > 1, 3,104 peaks open up in hypoxia and 2,506 open up in normoxia — hypoxia tends to open slightly more regions than it closes, but the response is clearly bidirectional rather than a uniform shift in one direction. This matches the broad, two-way chromatin remodeling reported by Dzwigonska et al. (2025) for hypoxic BV2-glioma co-cultures.
