---
title: "03 GO Biological Process Enrichment"
author: "Julia"
date: "2026-06-23"
output: html_document
---

# Introduction

Using the differentially accessible regions identified in `02_differential_accessibility.Rmd`,
we perform GO Biological Process enrichment to answer: **which biological processes
are more or less active in hypoxic versus normoxic glioma-associated microglia?**

More open chromatin in hypoxia → increased regulatory activity → upregulated process.  
More open chromatin in normoxia → decreased regulatory activity in hypoxia → downregulated process.

**Requires:** `01_downloading_data.Rmd` and `02_differential_accessibility.Rmd` run first.

---

```{r setup, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(SummarizedExperiment)
  library(rtracklayer)
  library(rGREAT)
  library(ggplot2)
})
```

---

# Step 1 — Load results from code 02

```{r load-results}
saved <- readRDS("results_da/part1_results.rds")
res <- saved$res    # edgeR topTags data.frame: logFC, FDR, etc.
se  <- saved$se     # SummarizedExperiment with all tested peaks as rowRanges
```

---

# Step 2 — Reconstruct peak coordinates

The `se` object from code 02 carries all tested peak coordinates in `rowRanges(se)`.
This is our background (all accessible chromatin tested) for GREAT.

```{r peak-coords}
# All tested peaks = proper background for GREAT
bg_regions <- rowRanges(se)
cat("Background (all tested peaks):", length(bg_regions), "\n")

# DA results are ordered by FDR; row order matches rowRanges(se) after edgeR filter.
# Re-attach coordinates to the results table.
# edgeR keeps the same row order as the input se after filterByExpr,
# so we match by name if available, otherwise by position.
res$coord_idx <- match(rownames(res), names(bg_regions))

# Peaks more open in hypoxia (logFC > 1, FDR < 0.05)
hyp_idx  <- which(!is.na(res$FDR) & res$FDR < 0.05 & res$logFC >  1)
# Peaks more open in normoxia (logFC < -1, FDR < 0.05)
norm_idx <- which(!is.na(res$FDR) & res$FDR < 0.05 & res$logFC < -1)

hyp_regions  <- bg_regions[hyp_idx]
norm_regions <- bg_regions[norm_idx]

cat("More open in hypoxia:",  length(hyp_regions),  "peaks\n")
cat("More open in normoxia:", length(norm_regions), "peaks\n")
```

---

# Step 3 — GO enrichment: processes INCREASED in hypoxia

```{r}
res_hyp <- great(
  hyp_regions,
  gene_sets  = "GO:BP",
  tss_source = "mm10",      # BV2 = mouse
  background = bg_regions,  # all tested peaks
  cores      = 2
)
bp_hyp_filt <- getEnrichmentTables(res_hyp)
cat("BP terms (increased in hypoxia):", nrow(bp_hyp_filt), "\n")
```

# Step 4 — GO enrichment: processes DECREASED in hypoxia

```{r}
res_norm <- great(
  norm_regions,
  gene_sets  = "GO:BP",
  tss_source = "mm10",
  background = bg_regions,
  cores      = 2
)
bp_norm_filt <- getEnrichmentTables(res_norm)
cat("BP terms (decreased in hypoxia):", nrow(bp_norm_filt), "\n")
```

# Step 5 — Plot (identical style to reference)

```{r}
# Increased in hypoxia
ggplot(head(bp_hyp_filt, 15),
       aes(fold_enrichment,
           reorder(description, p_adjust),
           size  = observed_region_hits,
           color = -log10(p_adjust))) +
  geom_point() +
  scale_color_viridis_c() +
  labs(title = "GO BP — increased in hypoxia", x = "Fold enrichment", y = NULL)

# Decreased in hypoxia
ggplot(head(bp_norm_filt, 15),
       aes(fold_enrichment,
           reorder(description, p_adjust),
           size  = observed_region_hits,
           color = -log10(p_adjust))) +
  geom_point() +
  scale_color_viridis_c() +
  labs(title = "GO BP — decreased in hypoxia", x = "Fold enrichment", y = NULL)
```

# Step 6 — Print results
```{r}
cat("\n=== PROCESSES INCREASED IN HYPOXIA ===\n")
print(head(bp_hyp_filt[, c("description", "observed_region_hits",
                             "fold_enrichment", "p_adjust")], 20))

cat("\n=== PROCESSES DECREASED IN HYPOXIA ===\n")
print(head(bp_norm_filt[, c("description", "observed_region_hits",
                              "fold_enrichment", "p_adjust")], 20))
```

