---
title: "02_differential accessibility"
output: html_document
date: "2026-06-24"
---

# Motif accessibility analysis
To identify motifs that are differentially acessible in between males and females in stressed vs control we do the following:  
1. Annotate peaks with GC bias  
2. Get a list of motifs  
3. Scan the ATAC seq peaks for motif matches  
4. Calculate the motif accessability deviations  
5. Perform differential analysis

## Setup
```{r}
suppressPackageStartupMessages({
  library(motifmatchr)
  library(MotifDb)
  library(universalmotif)
  library(ggplot2)
  library(ggrepel)
  library(sechm)
  library(GenomeInfoDb)
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(limma)
  library(betterChromVAR)
})
```

## Loading the count matrix
```{r}
se <- readRDS("data/peakCounts.rds")
colData(se)
```

I had an issue with running betterChromVAR without the following command. I remove peaks that have a zero value in every sample, because they contain no signal and can cause the background covariance calculation to fail.

```{r remove-zero}
non_zero <- Matrix::rowSums(SummarizedExperiment::assay(se)) > 0
se <- se[non_zero, ]
```

## Annotate regions
```{r annotation}
se <- GenomeInfoDb::keepStandardChromosomes(se, pruning.mode = "coarse")
seqlevelsStyle(se) <- "UCSC"
se <- addGCBias(se, genome=BSgenome.Mmusculus.UCSC.mm10)
```

## Motif matches
I will use the motif list from MotifDb. 
```{r motif-list}
motifs <- MotifDb::query(MotifDb, "HOCOMOCOv13")
```

Then we convert the motifs that they get accepted from motifmatchr.

```{r, message=FALSE}
motifs <- do.call(TFBSTools::PWMatrixList, convert_motifs(motifs, class="TFBSTools-PWMatrix"))
names(motifs) <- make.unique(sapply(motifs, \(x) gsub("\\..+","",x@name)))
```

And now we will scan the ATAC peaks for motif matches.

```{r motif-matches}
matches <- motifmatchr::matchMotifs(motifs, se, genome=BSgenome.Mmusculus.UCSC.mm10)
matches
```

## Get motif deviations
Now we will get the motif deviations, showing the realtive accessibility of each motif in each sample.

```{r motif-deviations}
dev <- betterChromVAR(se, matches)
dev
```

The output is also a SummarizedExperiment, this time with samples as columns and motifs as rows.
The values represent the relative accessibility of each motif in each sample.

# Differenial Analysis

We can use these motif deviations to perform differential analysis:


```{r}
dev$condition <- factor(dev$condition, levels = c("normoxia", "hypoxia"))
```

```{r DA}
design <- model.matrix(~condition, data=as.data.frame(colData(dev)))

fit <- eBayes(lmFit(assay(dev, "z"), design))
res <- as.data.frame(topTable(fit, number = Inf))
res$motif <- row.names(res)
```

```{r volcano-plot-DA}
res$diffexpressed <- "Not significant"
res$diffexpressed[res$logFC > 0.6 & res$adj.P.Val < 0.05] <- "Upregulated"
res$diffexpressed[res$logFC < -0.6 & res$adj.P.Val < 0.05] <- "Downregulated"

ggplot(data = res, aes(x = logFC, y = -log10(adj.P.Val), col = diffexpressed)) +
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
  geom_point(size = 2) + 
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  theme_classic()
```

```{r}
# Color classification column
res$diffexpressed <- "Not significant"
res$diffexpressed[res$logFC > 0.6 & res$adj.P.Val < 0.05] <- "Upregulated"
res$diffexpressed[res$logFC < -0.6 & res$adj.P.Val < 0.05] <- "Downregulated"

# Get the 10 up- and downregulated motifs for labelling
res$my_labels <- NA

top_up <- res[res$diffexpressed == "Upregulated", ]
top_up <- top_up[order(top_up$adj.P.Val), ]
top_up_motifs <- head(top_up$motif, 10)

top_down <- res[res$diffexpressed == "Downregulated", ]
top_down <- top_down[order(top_down$adj.P.Val), ]
top_down_motifs <- head(top_down$motif, 10)

target_motifs <- c(top_up_motifs, top_down_motifs)

# label column only for those target motifs
res$my_labels[res$motif %in% target_motifs] <- as.character(res$motif[res$motif %in% target_motifs])

ggplot(data = res, aes(x = logFC, y = -log10(adj.P.Val), col = diffexpressed, label = my_labels)) +
  geom_vline(xintercept = c(-0.6, 0.6), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
  geom_point(size = 2) + 
  scale_color_manual(values = c("#00AFBB", "grey", "#bb0c00"), 
                     labels = c("Downregulated", "Not significant", "Upregulated")) +
  theme_classic() +

  geom_text_repel(max.overlaps = Inf, show.legend = FALSE)
```

`normoxia` is the reference level so:  
- logFC > 0 means enrichment in hypoxia
- logFC < 0 means enrichment in normoxia


```{r}
col_rdbu <- hcl.colors(29, palette = "RdBu")
col_conditions <- list(
  condition = c("normoxia" = "#4682B4", "hypoxia" = "#b2182b")
)

sechm(dev, head(row.names(res)), 
      gaps_at = "condition", 
      assayName = "z",
      hmcols = cols_red_yellow_blue)
```