---
title: "The effect of hypoxic stress on glioma-associated myeloid cells"
author: "Olga Ledeneva, Yara Roth, Julia Shea"
date: "2026-07-03"
output: bookdown::html_document2
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ieee.csl
link-citations: true
---

# Introduction
Glioblastoma is one of the most common and aggressive malignancies, accounting for 50% of all malignant central nervous system tumors [@grochans_epidemiology_2022].
Despite the use of multimodal treatmets including surgery, radiotherapy and chemotherapy, the average survival rate of patients is only about 15 months [@grochans_epidemiology_2022].  
Ongoing challenges to glioblastoma treatment include its incomplete resection, high degree of heterogeneity and most importantly the immunosuppressive tumor microenvironment (TME) [@wu_glioblastoma_2021].
Tumor-associated myeloid cells, including macrophages and microglia cells are the most abundant population of immune, accounting up to 50% of the whole glioblastoma tumor mass [@khan_macrophages_2023]. Emerging evidence shows that these cells are critical components that promote tumor progression and induce the immunosuppressive TME [@khan_macrophages_2023]. Better understanding this the myloid cell population and it's role in the tumor microenvironment could help to resolve the complexity of the tumor and lead to novel therapeutic targets.  
Because glioblastomas contain highly proliferative cells, they display severely hypoxic regions [@bayona_tumor_2026]. Hypoxia has already been shown to drive glioma-associated myeloid cells toward an immunosuppressive phenotype and dysregulate their baseline identity [@sattiraju_hypoxic_2023].  
Therefore, the primary paper aimed to unvocer how hypoxia remodels chromatin accessiblity and the underlying epigenetic changes that control this pro-tumorigenic reprogramming of glioma-associated myeloid cells [@dzwigonska_hypoxic_2025].  

We worked with data from microglia-glioma co-cultures that were exposed to a hypoxic environment and compared them to ones cultured under normoxia.
To capture global epigenetic changes, the authors performed two high-throughput sequencing methods. First, they used ATAC-seq to profile changes in chromatin accessibility, and secondly, they performed CUT&RUN sequencing to map the H3K27ac histone modification, which marks active promoter and enhancer regions.  

Our computational analysis was structured into four major components to comprehensively evaluate the epigenomic effects of hypoxia:  

1. **Differential accessibility analysis** was performed to identify regions of chromatin that underwent significant structural changes upon hypoxic exposure.

2. **Functional enrichment analysis** was performed on these differentially accessible regions to uncover the overrepresented biological processes affected by hypoxia.

3. **Motif enrichment analysis** was performed to determine which transcription factor binding sites might be affected by hypoxia

4. **Integration of CUT&RUN data** was used to evaluate how closely active H3K27ac histone modification signals aligned with the observed changes in chromatin accessibility.

# Analysis
## Setup
First we will create a global plotting style.

```{r setup-plot, include=FALSE}
# knitr::opts_chunk$set(
#   fig.width = 8,
#   fig.height = 5,
#   fig.align = "center",
#   out.width = "90%"
# )
```

```{r}
# theme_set(
#   theme_classic() +
#   theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5)) #hjust = 0 to left align
#   )
```


## 01 - Downloading

## 02 - Differential Gene Accessibility

## 03 - Functional Enrichment Analysis

## 04 - Motif Enrichment Analysis

# Conclusion

# References